---
title: 'Module 1: Introduction to Unix and R'
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Patrick J Biggs, July 2020"
date: "15/07/2020"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Practical 1: A practical introduction to R

_In this lab session, you will experience use of R within the RStudio environment as an introduction to the main analyses we will perform for Modules 2 and 3 in the next part of the course._


### Introduction

As a brief summary of the lecture, we learnt that the statistical language R, its history, and that it has to be installed on a computer, but it is available for all the major platforms (Windows, OS X and Linux).  We talked about its use in this course as an environment in which to explore two aspects of bioinformatics – transcriptomics and metagenomics – in Modules 2 and 3 respectively.

We discussed the extensive help and support that these around R and its use, all the packages that can be downloaded to help you.  We then talked about the interface to R through the terminal (like last week), and then with a GUI version, and finally with full-blown environments such as RStudio, through which we also had a quick overview.  It is through the use of RStudio, that we will learn how to use R, and that is the focus of today’s practical.  

We looked at getting data in and out of R, its question and answer model of operation, and then at the five data types within R.  After looking at those, we went on to data formats, again a major focus of today’s lab practical.  We also touched on the installation of packages either through R itself for general package installation, as well as with Bioconductor for more specific bioinformatics applications.

So, now on to the practical…

First thing, as before, if you have an electronic copy of this practical, don’t copy and paste in the practical today! Why?  Typing these commands is part of the hard wiring required to make them stick! By typing commands it gives you another opportunity to think about the command before you execute it and typing is more likely to trigger a big red flag in dangerous situations!  Once again, as before, think twice, type once.

**This module is an introduction to the worlds of Linux and R, with a mastery test as an assessment.  Use this tutorial as a learning experience for a new set of tools, which in this case might go beyond this course, and into future study.**



### Conventions used for this RStudio practical

In what follows, we shall use the following typographical conventions for working with R:

•	Characters written in bold Consolas font are commands to be typed into the computer as they stand.  In a difference to practical 1, this will be for both functions and filenames, and all these commands should be typed in for this afternoon’s practical.
•	Characters written in this green italic Consolas font (e.g. “###”) indicate comments within the code that will help you.
•	Characters written in this dark red Consolas font (e.g. lines starting with “[1]”) indicate responses back from the R console inside RStudio that will help you too.  These will appear as light grey on the handouts but are in colour online.
•	Words inserted within square brackets [Enter] indicate keys to be pressed.

So, for example,

`> x <- c(1, 5, 7, 14)` [Enter]	- {NOTE: do not do this, it is an example.}


means "at the R console prompt ‘>’, type the code as written to make a vector x made up of the values 1, 5, 7 and 14, then press the key marked Enter".

Don't forget to press the `[Enter]` key: commands are not sent to the R console until this is done.  There will not be `[Enter]` at the end of every line of code, it is taken that this is to be typed at the end of the line unless otherwise stated.  It is also assumed that the first `‘>’` is the R console prompt.



### Opening up RStudio

You should have an RStudio icon on your machine, or you will have to go start one.  For us in this course, we have an icon on the application bar at the base of the screen.  It’s the blue circle with “R” in white writing inside it.  On the icon, just double click on it.  If you want to start one another way, click on the search magnifier icon in the top right of the screen, and type “RStudio”.  Either way, the result is a session of RStudio opening.  As you remember, our environment mostly looks like the image below:



Of course, your version will probably be looking a whole load emptier, as this is a screenshot taken when I was doing some work!  

We are going to be using RStudio today as a convenient introduction in using R, so we are going to focus mostly on the console today (bottom left).  However, we will be using – or rather you will need to be aware of – the whole environment for this afternoon, and also for Modules 2 and 3.



### Basic operations 

#### Assigning variables
Let’s start off by making the assigning the variable x with a value of 1:

``` {r assign1}
x = 1
x
```

We can also use the <- operator to do the same thing:
``` {r assign2}
x = 1
x
```

I prefer the “<-” as to me it is more demonstrative of what is going on i.e. we are making assigning x the value of 1.  Using “=” suggests equivalence, which is not strictly true.

#### Functions
We have seen this in the code conventions above.  Let’s repeat that, but this time call it `vec`:

``` {r func1}
vec <- c(1, 5, 7, 14)
vec
```

#### Comments
We have also seen this in the code conventions above.  Say we want to make a comment on some addition:
```{r comment}
1 + 1   ### this is a comment
```

This is our first example of using the 3 colours for this practical, and for all subsequent work within R.  In the console the comments will be the same colour, but if you wrote this in the script box within RStudio, you would see the overleaf, which are screenshots from my session.  This is due to something called syntax highlighting, which is not relevant for today (but useful nonetheless).  Of course, the script does not give you an answer, that is what the console is for.  
